Dynamic Space Efficient Cuckoo Table -- DySECT when doubling is too much!

### 2017-04-16 paper/paper.tex
Laptop - Paper Changes

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Dynamic Space Efficient Cuckoo Table when doubling is too much -- a novel approach to growing!}
\titlerunning{A Sample LIPIcs Article} %optional, in case that the title is too long; the running title should fit into the top page column

@@

\begin{abstract}
A lot of previous work concentrates on optimizing the speed and space
efficiency of hash tables.  Additionally many practical
implementations exist.  Modern open addressing techniques can fill a
given table to more than 95\% before the performance deteriorates
significantly.  But once the threshold is met, copying and migrating
the whole table still seems to be the most efficient answer.

To achieve any kind of efficiency, each such migration must be
amortized by newly inserted elements.  Therefore, each migration has
to increase the capacity by a significant amount -- usually, this
means doubling the capacity.  Growing by large factor
$\gamma$ often leads to bad space efficiency because the newly grown table
is only filled to less than a factor of $\gamma^{-1}$.

@@

One aspect which a lot of research focuses on is space efficiency.
Modern space efficient hash tables work well up to loads of 98\% and
more. But for this to matter, programmers have to know similarly tight
bounds to the absolute number of elements inserted into the table.
This is typically not the case, one of the most ubiquitous use cases
for hash tables is to unify data according to some key member.  To
guarantee good performance, we have to overestimate the necessary
capacity whenever the exact number of unique keys is not known a
priori.  Dynamic data-structures are necessary to alleviate this
problem.  In circumstances where space efficiency is important, space
efficiently growing hash tables are necessary.

@@

To avoid both of these pitfalls we propose a variant of bucket cuckoo
hashing (TODO REFERENCE).  Bucket cuckoo hashing is a technique where
each element can be stored in one of several associated constant sized
buckets.  When all of them are full we move elements, to create new
space.  We split one bucket cuckoo table into multiple subtables each
of which can grow independent from all others.  Because the buckets
associated with one element are spread over the different subtables,
growing one subtable alleviates pressure from all other tables.

Doubling the size of one subtable increases the overall size by a
small factor, while only moving a small number of elements -- making
it possible, to amortize these small size changes. The introduced size
and occupancy imbalance between subtables is alleviated using
displacement techniques common to cuckoo hashing. Using these
techniques, makes our table work and grow efficiently with fill
degrees around 97\% (TODO READ APPROPRIATE VALUE) .  Being
approximately \ldots{} \% faster than common hash table techniques under the
same circumstances.

\section{Motivation}
Space efficiency is a major factor for the development of new hashing
techniques.  The goal has long been to make hash tables work better in
high load scenarios.  Modern hashing techniques like Hopscotch, Robin
Hood, and Cuckoo Hashing can easily fill hash tables to more than 90\%
and up to 98\% while still achieving constant average running times on
most of their operations.  But since these techniques depend on
statically sized tables this only makes sense, if the number of unique
elements can be estimated tightly.

To visualize this assume the following scenario.  During the operation
of a word count benchmark, we know an upper bound $c$ to the number of
unique words. Therefore, we construct a table large enough to hold
these elements ($0.9^{-1}\cdot c$).  It is entirely possible that the
specific word count instance only contains $0.7\cdot c$ unique words.
The resulting table is therefore only filled 63\%.  It is clear that
better static sized tables cannot solve this issue.

Depending on the underlying problem it is impossible to tightly
estimate the final number of elements.  Dynamic growing can be a
solution even in these instances.  It is completely independent from
the problem's domain and thus also works in cases where estimating the
number of elements can be hard.  But to achieve space efficiency in
scenarios where the final size is not known the hash table has to grow
closely with the actual number of elements.  This is not efficiently
possible with any of the growing techniques used currently.

\section{Related Work}
\begin{itemize}
\item Some other engineering papers

\item Some Space efficient Stuff  (theoretical)

\item Some Space efficient Stuff  (practical)
\end{itemize}

\section{Preliminaries}
A hash table is a data-structure, which stores key-value-pairs
($\langle key, data \rangle$) and offers the following functions:
\verb~insert~ stores a given key-value pair, \verb~find~ given a key returns if
said key was stored, and if so it returns the stored value, and
\verb~erase~ which deletes a previously inserted element (if present).

@@

Throughout this paper, we use $n$ to denote the number of elements and
$m$ the number of cells ($m > n$) in any given hash table.  We define
the load factor as $load = n/m$.  Tables can usually only work
efficiently up to a certain load factor.  Above that, operations get
slower and/or have a possibility to fail.
When implementing a hash table one has to decide between storing
elements directly in the hash table -- \emph{Closed Hashing} -- or storing
pointers to elements -- \emph{Open Hashing}. This has an immediate impact
on the necessary memory.

@@

For large elements ($\gg size(pointer)$), one can simply use open
hashing, to reduce the relevant memory factor.  Therefore, we restrict
ourselves to the common case of elements whose size is comparable to
that of a pointer.  For our experiments we use 128bit elements (64bit
keys and 64bit values).  In these instances closed addressing
introduces a significant memory overhead (at least $\times1.5$),
therefore, we only consider closed hash tables for the purpose of this
paper. For closed hash tables, the memory efficiency is directly
dependent on the element density.

\subsection{$\alpha$-Space Efficient Hash Tables}
We call a hashing technique $\alpha$-space efficient when it can
work effectively using at most $\alpha \cdot size_{\min} + O(1)$
memory (TODO DEFINE WORKING EFFICIENTLY FURTHER i.e. constant average
running time or just look at experiments or inserting the last 10\%
elements is less than inserting all others). In this case we define
working efficiently as having average insertion times close to
$O(\frac{m}{m-n})$ (average number of probes to hit an empty cell).

@@

In many closed hashing techniques -- i.e. linear probing, cuckoo
hashing \ldots{} -- cells are the same size as elements. Therefore, being
$\alpha$ space efficient is the same as operating with a load factor
of $\alpha^{-1} = n/m$ ($\alpha\cdot n \cdot size(element) = m\cdot
size(element)$).  This is not necessarily the case for every kind of
hashing technique.  Some techniques like hopscotch hashing use
additional per cell information.  This would also be the case if we
store some per bucket data like the number of stored elements in our
bucket cuckoo hash table.

\subsubsection*{$\alpha$-Space Efficiency for Dynamic Tables}
The definition of a space efficient hashing technique given in (TODO
reference) the previous section is specifically targeted for
statically sized hash tables.  As detailed in Motivation (TODO
REFERENCE) our goal is to construct a hash table which can efficiently
(and tightly) grow with the number of inserted elements. Therefore, we
want to define the notion of an $\alpha$-space efficient dynamic
table.

@@

One point of discussion while defining dynamic space efficiency is the
space consumption during hash table operations.  During a table
migration both the source ($t_{source}$) and the target
($t_{target}$) table are allocated, thus using $mem =
(m_{source}+m_{target})\cdot size(cell)$ space.  This is especially
prohibiting if all elements are held in one table, then at least
$2\cdot m\cdot size(element)$ is necessary for each migration.

Our definition of dynamic space efficiency does not enforce size
reduction due to the deletion of elements, but it enforces that the
memory used by deleted elements must be reused (no tombstones).

\paragraph*{Trivial Examples}
One simple example of this is a linear probing table, which grows by a
factor of two, whenever 50\% of the table is filled. Here the table is
at least 25\% filled (immediately after growing), therefore, the table
is 6-space efficient (during the grow operation there are $(4+2)+n$
cells). Another example would be a cuckoo table which doubles in size
whenever it surpasses 95\%, fill rate. This would be a
$0.95^{-1}+0.45^{-1}\approx 3.2$-space efficient table.

\subsection{Cuckoo Hashing}
Cuckoo hashing is a technique to resolve hash conflicts in an closed
hashing hash table. Its main draw is that it guarantees constant
lookup times even in densely filled tables. The article TODO by TODO
(probably mitzenmacher with or without dietzfelbinger) et al. gives a
good overview over many cuckoo hashing variants.  The distinguishing
technique of cuckoo hashing is, that $H$ hash functions ($h_1, ... ,
h_H$) are used to find $H$ independent possible positions. Each
element is stored in one of its positions.  Even if all possible
positions are occupied one can often move elements to create space for
the current element. We call this displacing elements.

Bucket cuckoo hashing is a variant, where the cells of the hash table
are grouped into buckets of size $B$ ($m/B$ buckets).  Each element
appointed to one bucket can be stored in any of the bucket's cells .
Using buckets one can increase the number of displacement
opportunities drastically.

\textbf{Lookup:} Find and Remove operations have a guaranteed constant
running time. Independent from the tables density, there are $H$
possible buckets -- $H\cdot B$ cells -- that have to be searched to
find an element.

\textbf{Insert:} Each element is hashed to $H$ buckets.  If at least
one of those buckets has space left, then we store the element in the
bucket, that has the most space.  But when all buckets are full, then we
have to move elements around the table, such that a new space becomes
available.

To visualize the problem of displacing elements, one can think of the
directed graph implicitly defined by the hash table.  Each bucket
defines one node and each element defines pairwise connections between
the bucket it is stored in to its $H-1$ alternate buckets.  To insert
an element into the hash table we have to find a path from one of its
associated buckets, to a bucket, that has space remaining.  Then we
move elements along this path, to open a space on the initial bucket.
The two common techniques to find such paths are \emph{random walks} and
\emph{breadth first searches}.

\subsubsection*{Some Performance Bounds}
\begin{itemize}
\item bla how dense can we get
\end{itemize}

When randomly probing the table we expect to find a free cell after $\frac{1}{1-load factor}$

\section{DySECT Space Efficient Growing, by Splitting the Table}
Our goal is to build an efficiently growing data structure, that
remains space efficient at all times.  A commonly used growing
technique is to increase the size by migrating all elements
into a larger new table.  Of course this is
not memory efficient.  A natural thought which occurs is to double
only part of the data structure.  This idea together with the
balancing ability of multiple hash functions is what will achieve the
described functionality.

@@

Comparable to classic cuckoo hashing, each element has $H$ associated
buckets ($H$ number of hash functions), these can be in the same, or
in different subtables.  Each inserted element lies in one of its
associated buckets.  To find a bucket associated with an
element $e$, we compute $e$s hash value using the appropriated hash function
$h_i(e)$. We split that hash value into two parts, then we use the
first part to chose a subtable, and the second part to
choose one of the contained buckets.

@@

\subsection{Growing}
When the table contains enough elements, that the memory constraint
can be kept, we migrate a subtable.  We migrate subtables in order
from the first to the last, therefore, no subtable can be more than
twice as large as any other.

Assume that all subtables have $s=m/T$ cells. When $\alpha\cdot n >
m+2s$ We can grow the first subtable while keeping the size
constraint.  Doubling the size of a subtable increases the global
number of cells from $m_{old} = T\cdot s$ to $m_{new} = m_{old}+s =
(T+1)\cdot s$ (factor: $\frac{T+1}{T}$).  Note that all subsequent
grows migrate one of the smaller tables, until all tables have the
same size again.  Therefore, each grow until then increase the
capacity by the same absolute amount (smaller factor).

From a theoretical point of view, the cost of growing the subtable is
amortized by element insertions.  There are at least $\alpha^{-1}
\cdot s = \Omega(s)$ insertions between two migrations.  One migration
takes $\Theta(s)$ time. Alternatively abstract point of view, after
inserting $\alpha^{-1} \cdot s$ elements the table grows enough, to
raise the capacity about the same $\alpha^{-1} \cdot s$ effective
cells (actual cells $\times$ fill factor).

From a practical view, the migration is efficient because it accesses
cells in a linear fashion making it really cache efficient.  Even in
the target table cells are accessed linearly, because of the way we
assign elements to buckets there are no displacements necessary.  The
elements from each original bucket are split into two buckets of the
target table.  Therefore, no bucket of the target table can have more
elements than its respective original bucket.

In the implicit graph model of the cuckoo table (TODOsee section
cuckoo hashing), growing a subtable is equivalent to
splitting each node that represents a bucket within that table. The
edges (elements) in the implicitly defined subgraph are not doubled,
therefore, the resulting subgraph becomes more sparse, making it
easier to insert elements.

@@

When triggering the size reduction, one has to make sure that the
migration is amortized. Therefore, a grow operation cannot immediately
follow a shrink operation.  When shrinking is enabled we propose to
shrink one subtable when $\alpha*n < m-s'$ elements ($s'$ size of a
large table, $m_{new} = m_{old} -s'/2$).

Notice that the memory requirement actually increases during the
operation itself.
\subsection{Hashing Elements to Buckets}
\subsubsection*{Compute the Bucket from a Hashed Key}
From a hashed key, we have to compute the subtable and the bucket
within that subtable.  To make this efficient we use powers of two for
the number of subtables ($T = 2^t$), as well as the individual
subtable sizes ($s = 2^*$).  Since the number of subtables remains
constant we can use the first $t$ bits from the hashed key, to find the
appropriate subtable.  From the remaining bits we compute the offset
into this subtable using a bitmask ($\texttt{AND}~s-1 \leftrightarrow
\mod s$).

\subsubsection*{Reducing Number Computed Hash Functions}
Evaluating hash functions is expensive, therefore, reducing the number
of hash functions computed per operation increases the efficiency of
the table.  The hash function we use computes 64bit hash values
(i.e. xxHash TODO).  We split these 64bit value into two 32bit hash
values. All common bucket hash table sizes can be addressed using 32
bits. They can address up to $2^{32}$ buckets which achieves a
capacity for $2^{35} \approx 34$ billion elements (bucket size 8),
at that point a hash table consumes 512GiB memory.

@@

@@

For two hash functions ($H=2$) and only one grown table ($j=1$) this
means that $\approx 2n/(T+1)$ should be stored in the first table.
These are nearly all elements associated to a bucket in the first
table ($\approx 2n/T$). So to distribute elements evenly nearly all
elements would have to be stored there.

@@

Assume the global table is filled close to 100\% when the first table begins
to grow.  now there is the capacity for $s$ new elements, but this
capacity is only in the first table, elements that are not hashed to
the first table, automatically trigger displacements leading to slow
insertions.

Notice that repeated operations help to equalize this imbalance,
because elements are more likely inserted into the less dense areas,
and more likely to be deleted from more dense areas.

@@

\section{Experiments}
As described in section (TODO REFERENCE DIFFICULTIES\ldots{}) there are
many factors that impact hash table performance.  Many of these
factors are inherent to the combination of growing and space
efficiency.  To show that our ideas work from a practical point of
view we use a wide portfolio of both micro-benchmarks and practical
experiments.

@@

\subsubsection*{Choice of Constants}
In our tests, we use three hash functions ($H=3$) and a bucket size of
($B=8$). These values have consistently outperformed other
possibilities.  We chose to use 256 subtables for all our tests
($T=256$).  The number of subtables only plays a minor role for the
running time.

\subsubsection*{Comparison Implementations}
To put the performance of our solution into perspective, we
implemented and tested several other options for space efficient
hashing.  We use our own implementations, since most hash tables found
online do not support strict space-efficiency.  In Appendix TODO we
show that our implementations are at least competitive with
implementations found .

@@

To grow our linear probing table, we use table migrations.  This can
be made space efficient, by increasing the capacity in place.  This is
possible due to over allocation (MAYBE SOME SOURCE/APPENDIX/WRITE
SHORT PAPER) and a simple in place migration algorithm (ALSO
APPENDIX/SHORT PAPER).  For a table that was initialized to hold $n'$
elements with a space-efficiency $\alpha$ we trigger growing once the
number of elements is $n = \frac{(1+\alpha)}{2}n'$. We then increase
the capacity $m$ to $\alpha\cdot n$ to remain strictly size-efficient.
Our implementation of the migration is highly cache efficient, but
all elements have to be moved.  This makes repeated migrations inefficient
especially for small $\alpha$.

@@

\paragraph*{Robin Hood Hashing} is another technique based on linear probing. The
goal of this technique is to improve find performance by minimizing
the maximum/average displacement distance.  The drawback of this technique is
that insertions have to compute the hash of each element that is
skipped by the insertion.

\paragraph*{Classic Bucket Cuckoo Hashing} is the technique we adopted to
design our hashing technique.  It has clear advantages for space
efficient tables.  Like our variant it offers constant lookup and
deletion times.  In contrast to our variant, it does not suffer from
any imbalances caused either by table resolution or by local growing
operations.  Growing works similar to the linear probing table above.
Similar to DySECT, $H=3$ and $B=8$ have turned out to be an optimum
for performance.

@@

\subsection{Influence of Fill Ratio under a Static Table Size}
\label{sec:exp_eps}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{../eval/plots/eps_insert.pdf}
  \caption{\label{fig:eps_insert}Insert timings at different fill
    degrees into a non-growing table.}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.49\textwidth]{../eval/plots/eps_find_s.pdf}
  \includegraphics[width=0.49\textwidth]{../eval/plots/eps_find_u.pdf}
  \caption{Timings for find operations with different fill degrees
    (\emph{left:}successful, \emph{right:}unsuccessful).}
  \label{fig:eps_find}
\end{figure}
Our hash table is designed for space efficient growing, therefore, its
main advantages show in dynamic workloads.  We still performed some
tests with a static table size.  With these measurements, we can
highlight different points without the influence of dynamically growing.

>>>>

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Dynamic Space Efficient Cuckoo Table -- DySECT when doubling is too much!} % -- a novel approach to growing!}
\titlerunning{A Sample LIPIcs Article} %optional, in case that the title is too long; the running title should fit into the top page column

@@

\begin{abstract}
Research on hash table data-structures has concentrated on optimizing
the speed and space efficiency for a long time.  Modern open
addressing techniques can fill a given table to more than 95\% before
the performance deteriorates significantly.  But once the threshold is
met, copying and migrating the whole table still seems to be the most
efficient answer.

To achieve any kind of efficiency, each such migration must be
amortized by newly inserted elements.  Therefore, each migration has
to increase the capacity by a significant amount -- usually, this
means doubling the capacity.  Growing by a large factor $\gamma$ leads
to bad space efficiency because the newly grown table is only filled
less than a factor of $\gamma^{-1}$.

@@

One aspect which a lot of research focuses on is space efficiency.
Modern space efficient hash tables work well up to loads of 95\% and
more. But for this to matter, programmers have to know similarly tight
bounds to the absolute number of elements inserted into the table.
This is typically not the case, one of the most ubiquitous use cases
for hash tables is to unify data elements by their key.  To
guarantee good performance, we have to overestimate the necessary
capacity whenever the exact number of unique keys is not known a
priori.  Dynamic data-structures are necessary to alleviate this
problem.  In circumstances where space efficiency is important, space
efficiently growing hash tables are necessary.

@@

To avoid both of these pitfalls we propose a variant of bucket cuckoo
hashing (TODO REFERENCE).  Bucket cuckoo hashing is a technique where
each element can be stored in one of several associated constant sized
buckets.  When all of them are full we move elements, to create new
space.  To solve the problem of efficient migration, we split one
bucket cuckoo table into multiple subtables each of which can grow
independently from all others.  Because the buckets associated with
one element are spread over the different subtables, growing one
subtable alleviates pressure from all others by allowing moves from a
dense table to the grown table.

Doubling the size of one subtable increases the overall size by a
small factor, while only moving a small number of elements -- making
the size changes easy to amortize. The introduced size and occupancy
imbalance between subtables is alleviated using displacement
techniques common to cuckoo hashing. Using these techniques, makes our
table work and grow efficiently with fill degrees of 95\% and more.
Being approximately TODO \ldots{} \% faster than common hash table
techniques under the same circumstances.

\section{Motivation}
\label{sec:motication}
Space efficiency is a major factor for the development of new hashing
techniques.  The goal has long been to make hash tables work better in
high load scenarios.  Modern hashing techniques like Hopscotch, Robin
Hood, and Cuckoo Hashing can easily fill hash tables to more than 90\%
and up to 98\% while still achieving constant average running times on
most of their operations.  But since these techniques depend on
statically sized tables this only makes sense, if the number of unique
elements can be estimated with similar tight bounds.

To visualize this assume the following scenario.  During the operation
of a word count benchmark, we know an upper bound $c$ to the number of
unique words. Therefore, we construct a table large enough to hold
these elements ($0.9^{-1}\cdot c$ cells).  It is entirely possible
that the specific word count instance only contains $0.7\cdot c$
unique words (or less).  The resulting table is therefore only filled
63\%.  It is clear that better static sized tables cannot solve this
issue.

Depending on the underlying problem it is impossible to tightly
estimate the final number of elements.  Dynamic growing can be a
solution even in these instances.  It is completely independent from
the problem's domain and thus also works in cases where estimating the
number of elements can be hard.  To achieve space efficiency in
scenarios where the final size is not known the hash table has to grow
closely with the actual number of elements.  This is not efficient
with any of the current techniques used for hashing and mgration.

\section{Related Work}
As previously stated, there has been a lot of research on hash tables,
since they were introduced in the 1950s.

\section{Preliminaries}
A hash table is a data-structure, which stores key-value-pairs
($\langle key, data \rangle$) and offers the following functions:
\verb~insert~ -- stores a given key-value pair; \verb~find~ -- given a key returns if
said key was stored, and if so it returns the stored value; and
\verb~erase~ -- deletes a previously inserted element (if present).

@@

Throughout this paper, we use $n$ to denote the number of elements and
$m$ the number of cells ($m > n$) in any given hash table.  We define
the load factor as $load = n/m$, and we use $\epsilon = 1-load$ for
the fraction of available cells.  Tables can usually only work
efficiently up to a certain load factor.  Above that, operations get
slower and/or have a possibility to fail.  When implementing a hash
table one has to decide between storing elements directly in the hash
table -- \emph{Closed Hashing} -- or storing pointers to elements --
\emph{Open Hashing}. This has an immediate impact on the necessary
memory.

@@

For large elements ($\gg size(pointer)$), one can simply use open
hashing, to reduce the relevant memory factor.  Therefore, we restrict
ourselves to the common case of elements whose size is comparable to
that of a pointer.  For our experiments we use 128bit elements (64bit
keys and 64bit values).  In these instances closed addressing
introduces a significant memory overhead (at least $\times1.5$),
therefore, we only consider closed hash tables for the purpose of this
paper. For closed hash tables, the memory efficiency is directly
dependent on the element density.  To reach high fill degrees with
closed hashing tables, we have to employ \emph{open addressing}
techniques which means, that elements are not stored in predetermined
cells, but can be stored in one of several options (e.g. linear
probing, or cuckoo hashing).

\subsection{$\alpha$-Space Efficient Hash Tables}
\label{sec:pre_staticspace}
We call a hashing technique $\alpha$-space efficient when it can work
effectively using at most $\alpha \cdot size_{\min} + O(1)$ memory. In
this case we define working efficiently as having average insertion
times close to $O(\frac{1}{\varepsilon})$. This is a natural bound
since it is the expected number of fully random probes needed to hit
an empty cell.

@@

In many closed hashing techniques (e.g.~linear probing, cuckoo
hashing) cells are the same size as elements. Therefore, being
$\alpha$ space efficient is the same as operating with a load factor
of $\alpha^{-1} = n/m$ (because $\alpha\cdot n \cdot size(element) = m\cdot
size(element)$).  Because of this, we will mostly talk about the load
factor of a table instead of its memory usage.

Some techniques like hopscotch hashing use additional per cell
information.  These hash table architectures, have to be filled more
densely to achieve the same space efficiency.

\paragraph*{$\alpha$-Space Efficiency for Dynamic Tables}
The definition of a space efficient hashing technique given aboce is
specifically targeted for statically sized hash tables.  As detailed
in Section~\ref{sec:motivation} our goal is to construct a hash table
which can efficiently (and tightly) grow with the number of inserted
elements. Therefore, we want to define the notion of an $\alpha$-space
efficient dynamic table.

@@

One problem for many implementations of space efficient hash tables is
the migration.  During a normal full table migration, both the
original table, and the new table are allocated.  This takes
$\textit{m}_{\textit{new}} + \textit{m}_{\textit{old}}$
cells. Therefore, no normal full table migration can be more then
2-space efficient.  The only option to perform a full table migration
with less memory, is to increase the memory in place (see
Section~\ref{sec:dys_inplace} and
Section~\ref{sec:exp_competitor}).

Similar to static $\alpha$-space efficiency, we will mostly talk about
the \emph{minimum load factor}, instead of $\alpha$.  It should be
noted, that this definition does not enforce size reduction due to the
deletion of elements, but it enforces that the cells used by deleted
elements must be reused.

\subsection{Cuckoo Hashing}
Cuckoo hashing is a technique to resolve hash conflicts in a closed
hashing hash table. Its main draw is that it guarantees constant
lookup times even in densely filled tables. The article TODO by TODO
(probably mitzenmacher with or without dietzfelbinger) et al. gives a
good overview over many cuckoo hashing variants.  The distinguishing
technique of cuckoo hashing is, that $H$ hash functions ($h_1, ... ,
h_H$) are used to find $H$ independent possible positions. Each
element is stored in one of its positions.  Even if all possible
positions are occupied one can often move elements to create space for
the current element. We call this process displacing elements.

Bucket cuckoo hashing is a variant, where the cells of the hash table
are grouped into buckets of size $B$ ($m/B$ buckets).  Each element
appointed to one bucket can be stored in any of the bucket's cells .
Using buckets one can drastically increase the number of displacement
opportunities.

\emph{Find} and \emph{remove} operations have a guaranteed constant
running time. Independent from the tables density, there are $H$
possible buckets -- $H\cdot B$ cells -- that have to be searched to
find an element (elements can be removed easily).

During an \emph{insert} the element is hashed to $H$ buckets.  If at least
one of those buckets has space left, then we store the element in the
bucket, that has the most space.  But when all buckets are full, then we
have to move elements around the table, such that a new space becomes
available.

\label{sec:cuckoo_graph}
To visualize the problem of displacing elements, one can think of the
directed graph implicitly defined by the hash table.  Each bucket
defines one node and each element defines pairwise connections between
the bucket it is stored in to its $H-1$ alternate buckets.  To insert
an element into the hash table we have to find a path from one of its
associated buckets, to a bucket, that has space remaining.  Then we
move elements along this path, to open a space on the initial bucket.
The two common techniques to find such paths are \emph{random walks} and
\emph{breadth first searches}.

\begin{itemize}
\item copy some performance bounds from a cuckoo hashing paper
\item bla how dense can we get
\end{itemize}

When randomly probing the table we expect to find a free cell after
$\frac{1}{\varepsilon}$ random probes.

\section{DySECT Space Efficient Growing, by Splitting the Table}
Our goal is to build an efficiently growing data structure, that
remains space efficient at all times.  A commonly used growing
technique is to double the size by migrating all elements
into a larger new table.  Of course this is
not memory efficient.  A natural thought which occurs is to double
only part of the data structure.  This idea together with the
balancing ability of multiple hash functions is what will achieve the
described functionality.

@@

Comparable to classic cuckoo hashing, each element has $H$ associated
buckets ($H$ number of hash functions), these can be in the same, or
in different subtables.  Each inserted element lies in one of its
associated buckets.  To find a bucket associated with an element $e$,
we compute $e$'s hash value using the appropriate hash function
$h_i(e)$, from a hashed key, we have to compute the subtable and the bucket
within that subtable.

To make this efficient we use powers of two for the number of
subtables ($T = 2^t$), as well as the individual subtable sizes ($s =
2^x$).  Since the number of subtables remains constant we can use the
first $t$ bits from the hashed key, to find the appropriate subtable.
From the remaining bits we compute the offset into this subtable using
a bitmask ($h_i(e)~\texttt{AND}~s-1 \leftrightarrow h_i(e) \mod s$).

%% We split that hash value into two parts, then we use the
%% first part to chose a subtable, and the second part to
%% choose one of the contained buckets.

@@

\subsection{Growing}
We allocate grow one subtable,
when the table contains enough elements, that the memory constraint
can be kept during the migration.  We migrate subtables in order
from the first to the last, therefore, no subtable can be more than
twice as large as any other.

Assume that all subtables have $s=m/T$ cells. When $\alpha\cdot n >
m+2s$ we can grow the first subtable while keeping the size
constraint.  Doubling the size of a subtable increases the global
number of cells from $m_{old} = T\cdot s$ to $m_{new} = m_{old}+s =
(T+1)\cdot s$ (factor: $\frac{T+1}{T}$).  Note that all subsequent
grows migrate one of the smaller tables, until all tables have the
same size again.  Therefore, each grow until then increase the
capacity by the same absolute amount (smaller factor).

The cost of growing the subtable is amortized by all insertions that
happened since the last subtable migration.  There are at least
$\alpha^{-1} \cdot s = \Omega(s)$ insertions between two migrations.
One migration takes $\Theta(s)$ time.  Appart from being amortized,
the migration is efficient in practise because it accesses cells in a
linear fashion making it really cache efficient.  Even in the target
table cells are accessed linearly.  Because of the way we assign
elements to buckets, one original bucket is split into two target
buckets.  Therefore, no displacements are necessary, since no bucket
can be overfilled.

In the implicit graph model of the cuckoo table
(Section~\ref{sec:cuckoo_graph}), growing a subtable is equivalent to
splitting each node that represents a bucket within that table. The
edges (elements) in the implicitly defined subgraph are not doubled,
therefore, the resulting subgraph becomes more sparse, making it
easier to insert elements.

@@

When triggering the size reduction, one has to make sure that the
migration is amortized. Therefore, a grow operation cannot immediately
follow a shrink operation.  When shrinking is enabled we propose to
shrink one subtable when $\alpha\cdot n < m-s'$ elements ($s'$ size of a
large table, $m_{new} = m_{old} -s'/2$).

Notice that the allocated memory actually increases during the
migration itself.

\subsection{Implementation Details}
For our experiments in Section~\ref{sec:exp} we use three hash
functions ($H=3$) and a bucket size of ($B=8$). These values have
consistently outperformed other possibilities.  We chose to use
$T=256$ subtables for all our tests.  The number of subtables only
plays a minor role for the running time.  We use breadth first search
as displacement strategy. Breadth first search performs better in
bucket cuckoo hashing, since it better uses the read cache lines from
one bucket.

\subsubsection*{Reducing Number Computed Hash Functions}
Evaluating hash functions is expensive, therefore, reducing the number
of hash functions computed per operation increases the efficiency of
the table.  The hash function we use computes 64bit hash values
(i.e. xxHash TODO).  We split the 64bit value into two 32bit hash
values. All common bucket hash table sizes can be addressed using 32
bits. They can address up to $2^{32}$ buckets which achieves a
capacity for $2^{35} \approx 34$ billion elements (bucket size 8),
at that point a hash table consumes 512GiB memory.

@@

\subsubsection*{Table Architecture (two Variants)}
\label{sec:dys_inplace}
The hash table can be implemented as a collection of pointers to
subtables. We have to lookup the corresponding pointer, whenever a
subtable is accessed.  This does not impact performance much since all
subtable pointers will usually be cached.

Another option, which offers some benefits is to use virtual memory,
and memory overcommiting.  The idea is that the operating system will
usually allow larger memory allocations than the machine has memory,
with the anticipation that not all the allocated memory will actually
be used.  Only memory pages, that are actually used will be mapped
from virtual to physical memory pages.  Thus from a strict reasource
point of view, the memory is not yet used.  Therefore, we can also
implement the subtables as sections within one huge allocation.  This
has the advantage, that the offset for each table can be quickly
computed, without looking it up from a table.

The added advantage is that we can grow subtables in-place. To map
virtual to physical memory we have to initialize part of the table. To
increase the size of a subtable, it is enough to initialize a new
section of the table.  Once this is done, we have to redistribute all
elements between the two parts of the table (moving only half the
elements).  This allows us to grow a subtable without the space
overhead of reallocation.

This way, we can also implement dynamic space efficient variants of
several other hashing schemes (see Section~\ref{sec:exp_competitors}).
These implementations, can easily compete with similar implementations
that use reallocation.

@@

For two hash functions ($H=2$) and only one grown table ($j=1$) this
means that $\approx 2n/(T+1)$ should be stored in the first table, to
achieve a balanced bucket distribution.  Therefore, nearly all
elements associated to a bucket in the first table ($\approx 2n/T$)
have to be stored there.

@@

Assume the global table is filled close to 100\% when the first table
begins to grow.  now there is the capacity for $s$ new elements, but
this capacity is only in the first table, elements that are not hashed
to the first table, automatically trigger displacements leading to
slow insertions.  Notice that repeated operations help to equalize
this imbalance, because elements are more likely inserted into the
less dense areas, and more likely to be deleted from more dense areas.

@@

\section{Experiments}
\label{sec:exp}
There are many factors that impact hash table performance.  Many of
these factors are inherent to the combination of growing and space
efficiency.  To show that our ideas work in practice we use a
portfolio of both micro-benchmarks and practical experiments.

@@

\subsubsection*{Comparison Implementations}
\label{sec:exp_competitor}
To put the performance of our solution into perspective, we
implemented and tested several other options for space efficient
hashing.  We use our own implementations, since most hash tables found
online do not support strict space-efficiency.  In Appendix TODO we
show that our implementations are at least competitive with
implementations found .

@@

To grow our linear probing table, we use full table migrations.  This
can be made space efficient, by increasing the capacity in place.
This is possible due to over allocation (similar to
Section~\ref{sec:dys_inplace}) and a simple in place migration
algorithm (APPENDIX OR SECTION?).  For a table that was
initialized with $m$ cells with a min load factor $1-\varepsilon$ we
trigger growing once the table is loaded more than
$1-\frac{\varepsilon}{2}$. We then increase the capacity $m$ to
$(1-\varepsilon)^{-1}\cdot n$ to remain strictly size-efficient.  Our implementation
of the migration is highly cache efficient, but all elements have to
be moved.  This makes repeated migrations inefficient especially for
small $\alpha$.

@@

TODO DESCRIBE 2 CELL BUCKET VARIANT

\paragraph*{Robin Hood Hashing} is another technique based on linear probing. The
goal of this technique is to improve find performance by minimizing
the maximum/average displacement distance.  The drawback of this
technique is that insertions have to compute the hash of each element
that is skipped by the insertion.  Migrating a Robin Hood hash table
is very fast, since the additional hash computations are not
necessary, and the element order remains the same (sorted by hash value).

\paragraph*{Classic Bucket Cuckoo Hashing} is the technique we adopted to
design our hashing technique.  Like our variant it offers constant
lookup and deletion times.  But it suffers from an inefficient
growing technique similar to the one used by the linear probing table above.
Similar to DySECT, $H=3$ and $B=8$ have turned out to be an optimum
for performance.

@@

\subsection{Influence of Fill Ratio under a Static Table Size}
\label{sec:exp_eps}

The following test was performed, by initializing a table with
$m\approx20M$~cells (non-growing). Then elements are inserted, until there
is a critical amount of failing insertions ($>100$).  At different
stages, we measure the speed of new insertions
(Figure~\ref{fig:eps_insert}), and finds (Figure~\ref{fig:eps_find}).
Performance is measured by performing each operation 1000 times, finds
are measured using either randomly selected elements from within the table (successful), or by
searching random elements from key space (unsuccessful).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{../eval/plots/eps_insert.pdf}
  \caption{\label{fig:eps_insert}Insert timings at different fill
    degrees into a non-growing table.}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.49\textwidth]{../eval/plots/eps_find_s.pdf}
  \includegraphics[width=0.49\textwidth]{../eval/plots/eps_find_u.pdf}
  \caption{Timings for find operations with different fill degrees
    (\emph{left:}successful, \emph{right:}unsuccessful).}
  \label{fig:eps_find}
\end{figure}

As one might expect, the performance, of insert operations depends
highly on the fill degree of the table.  Therefore, we show it
normalized with $\frac{1}{\varepsilon}$, which is the expected number
of fully random probes, to find a free cell and thus a natural
estimate for the running time. We see that -- up to a certain point --
for all tables the insertion time behaves proportional to
$\frac{1}{\varepsilon}$, then it begins to decrease uncontrolled.  The
plot shows, that our table remains stable for longer than linear
probing variants, but not as long, as for classic cuckoo hashing.
Hopscotch hashing fails earlier than all other tables, because its
bitmaps become to small, to store the neighborhood information.
